---
title: "2nd Chapter"
author: "Mykola Dereva"
date: "1/11/2021"
output: html_document
---

```{r}
library(ggplot2)
library(tidyr)
library(ggeasy)
library(dplyr)
```



2E1. Which of the expressions below correspond to the statement: the probability of rain on Monday?

  (1) Pr(rain)
  (2) Pr(rain|Monday)
  (3) Pr(Monday|rain)
  (4) Pr(rain, Monday)/ Pr(Monday)

Answer: 2, 4

2E2. Which of the following statements corresponds to the expression: Pr(Monday|rain)?

  (1) The probability of rain on Monday.
  (2) The probability of rain, given that it is Monday.
  (3) The probability that it is Monday, given that it is raining.
  (4) The probability that it is Monday and that it is raining.

Answer: 3

2E3. Which of the expressions below correspond to the statement: the probability that it is Monday,
given that it is raining?

  (1) Pr(Monday|rain)
  (2) Pr(rain|Monday)
  (3) Pr(rain|Monday) Pr(Monday)
  (4) Pr(rain|Monday) Pr(Monday)/ Pr(rain)
  (5) Pr(Monday|rain) Pr(rain)/ Pr(Monday)

Answer: 1

2E4. The Bayesian statistician Bruno de Finetti (1906–1985) began his 1973 book on probability theory with the declaration: “PROBABILITY DOES NOT EXIST.” The capitals appeared in the original,
so I imagine de Finetti wanted us to shout this statement. What he meant is that probability is a device for describing uncertainty from the perspective of an observer with limited knowledge; it has no objective reality. Discuss the globe tossing example from the chapter, in light of this statement. What does it mean to say “the probability of water is 0.7”?

Answer: As for me, this means that even though we do not know the "true" proportion of the globe
covered with the water, but given the observations we have, our best guess is that there is a 70% chance of getting a "water" in the next toss of a globe.
In other words, we can estimate the continuous range of probabilities even if in each toss we
can get only a binary outcome (0 - land; 1 - water). 
This probability helps us to understand the proportion of a water on the Earth, however
it exists only in our imagination. 


2M1. Recall the globe tossing model from the chapter. Compute and plot the grid approximate
posterior distribution for each of the following sets of observations. In each case, assume a uniform prior for p.
  (1) W, W, W
  (2) W, W, W, L
  (3) L, W, W, L, W, W, W


```{r}
# define grid length 
grid_l <- 30

# define grid
p_grid <- seq(from = 0 , to = 1 , length.out = grid_l)

# define prior
prior <- rep(1, grid_l)

# compute likelihood at each value in grid
likelihood_1 <- dbinom(3, size = 3, prob = p_grid)
likelihood_2 <- dbinom(3, size = 4, prob = p_grid)
likelihood_3 <- dbinom(5, size = 7, prob = p_grid)

# compute product of likelihood and prior
unstd.posterior_1 <- likelihood_1 * prior
unstd.posterior_2 <- likelihood_2 * prior
unstd.posterior_3 <- likelihood_3 * prior

# standardize the posterior, so it sums to 1
posterior_1 <- unstd.posterior_1 / sum(unstd.posterior_1)
posterior_2 <- unstd.posterior_2 / sum(unstd.posterior_2)
posterior_3 <- unstd.posterior_3 / sum(unstd.posterior_3)
```

```{r}
# Plot a results
data.frame(p_grid, posterior_1, posterior_2, posterior_3) %>%
  pivot_longer(cols = starts_with("post"), names_to = "posterior", values_to = "prob") %>%
  mutate(posterior = case_when(posterior == "posterior_1" ~ "W, W, W",
                               posterior == "posterior_2" ~ "W, W, W, L",
                               posterior == "posterior_3" ~ "L, W, W, L, W, W, W")) %>% 
  ggplot(aes(x = p_grid, y = prob, col = posterior)) +
  geom_point() +
  geom_line() +
  labs(x = "Proportion of Water",
       y = "Posterior Probability",
       color = "Pattern",
       title = "Uniformal Priors") +
  theme_minimal()
```



2M2. Now assume a prior for p that is equal to zero when p < 0.5 and is a positive constant when
p ≥ 0.5. Again compute and plot the grid approximate posterior distribution for each of the sets of
observations in the problem just above

```{r}
# define grid length 
grid_l <- 30

# define grid
p_grid <- seq(from = 0 , to = 1 , length.out = grid_l)

# define new prior
prior <- ifelse(p_grid < 0.5, yes = 0, no = 1)

# compute likelihood at each value in grid
likelihood_1 <- dbinom(3, size = 3, prob = p_grid)
likelihood_2 <- dbinom(3, size = 4, prob = p_grid)
likelihood_3 <- dbinom(5, size = 7, prob = p_grid)

# compute product of likelihood and prior
unstd.posterior_1 <- likelihood_1 * prior
unstd.posterior_2 <- likelihood_2 * prior
unstd.posterior_3 <- likelihood_3 * prior

# standardize the posterior, so it sums to 1
posterior_1 <- unstd.posterior_1 / sum(unstd.posterior_1)
posterior_2 <- unstd.posterior_2 / sum(unstd.posterior_2)
posterior_3 <- unstd.posterior_3 / sum(unstd.posterior_3)


# Plot a results
data.frame(p_grid, posterior_1, posterior_2, posterior_3) %>%
  pivot_longer(cols = starts_with("post"), names_to = "posterior", values_to = "prob") %>%
  mutate(posterior = case_when(posterior == "posterior_1" ~ "W, W, W",
                               posterior == "posterior_2" ~ "W, W, W, L",
                               posterior == "posterior_3" ~ "L, W, W, L, W, W, W")) %>% 
  ggplot(aes(x = p_grid, y = prob, col = posterior)) +
  geom_point() +
  geom_line() +
  labs(x = "Proportion of Water",
       y = "Posterior Probability",
       color = "Pattern",
       title = "Priors = 0 when p < 0.5") +
  theme_minimal()
```



2M3. Suppose there are two globes, one for Earth and one for Mars. The Earth globe is 70% covered
in water. The Mars globe is 100% land. Further suppose that one of these globes—you don’t know
which—was tossed in the air and produced a “land” observation. Assume that each globe was equally
likely to be tossed. Show that the posterior probability that the globe was the Earth, conditional on seeing “land” (Pr(Earth|land)), is 0.23

```{r}
# Define probability of land on Earth and Mars
p_land_earth <- 0.3
p_land_mars <- 1

# Earth prior
p_earth <-  0.5

# average probability of land
p_land <- (p_land_earth * p_earth) + (p_land_mars * (1 - p_earth))

# Bayes theorem
p_earth_land <- (p_land_earth * p_earth) / p_land

print(p_earth_land)
```



2M4. Suppose you have a deck with only three cards. Each card has two sides, and each side is either black or white. One card has two black sides. The second card has one black and one white side. The third card has two white sides. Now suppose all three cards are placed in a bag and shuffled. Someone reaches into the bag and pulls out a card and places it flat on a table. A black side is shown facing up, but you don’t know the color of the side facing down. Show that the probability that the other side is also black is 2/3. 
Use the counting method (Section 2 of the chapter) to approach this problem. This means counting up the ways that each card could produce the observed data (a black side facing up on the table).

```{r}
ways_to_have_black <- c(2, 1, 0)

prior <- rep(1, 3)

posterior <- (ways_to_have_black * prior)
posterior <- posterior / sum(posterior)

# Only the first card have the second black side
# thus the p of first card and p of second black side is the same
p_black <- posterior[1]
print(posterior[1])
```


  2M5. Now suppose there are four cards: B/B, B/W, W/W, and another B/B. Again suppose a card is
drawn from the bag and a black side appears face up. Again calculate the probability that the other
side is black

```{r}
ways_to_have_black <- c(2, 1, 0, 2)

prior <- rep(1, 4)

posterior <- (ways_to_have_black * prior)
posterior <- posterior / sum(posterior)

p_black <- posterior[1] + posterior[4]
p_black
```



  2M6. Imagine that black ink is heavy, and so cards with black sides are heavier than cards with white sides. As a result, it’s less likely that a card with black sides is pulled from the bag. So again assume there are three cards: B/B, B/W, and W/W. After experimenting a number of times, you conclude that for every way to pull the B/B card from the bag, there are 2 ways to pull the B/W card and 3 ways to pull the W/W card. Again suppose that a card is pulled and a black side appears face up. Show that the probability the other side is black is now 0.5. Use the counting method, as before

```{r}
ways_to_have_black <- c(2, 1, 0)

prior <- c(1, 2, 3)

posterior <- (ways_to_have_black * prior)
posterior <- posterior / sum(posterior)

p_black <- posterior[1]
p_black
```



  2M7. Assume again the original card problem, with a single card showing a black side face up. Before looking at the other side, we draw another card from the bag and lay it face up on the table. The face that is shown on the new card is white. Show that the probability that the first card, the one showing a black side, has black on its other side is now 0.75. Use the counting method, if you can. 
  Hint: Treat this like the sequence of globe tosses, counting all the ways to see each observation, for each possible first card
  
```{r}
ways_to_have_first_black <- c(2, 1, 0)
ways_to_have_second_white <- c(3, 2, 1)

prior <- c(1, 1, 1)

posterior <- (ways_to_have_first_black * ways_to_have_second_white * prior)
posterior <- posterior / sum(posterior)

p_black <- posterior[1]
p_black
```
  
  2H1. Suppose there are two species of panda bear. Both are equally common in the wild and live
in the same places. They look exactly alike and eat the same food, and there is yet no genetic assay capable of telling them apart. They differ however in their family sizes. Species A gives birth to twins 10% of the time, otherwise birthing a single infant. Species B births twins 20% of the time, otherwise birthing singleton infants. Assume these numbers are known with certainty, from many years of field research.
  Now suppose you are managing a captive panda breeding program. You have a new female panda
of unknown species, and she has just given birth to twins. What is the probability that her next birth will also be twins?

```{r}
# Define A and B species probability of getting twins
p_twins <- c(0.1, 0.2)

priors <- c(1, 1)

p_first_twin <- (p_twins * priors) / sum(p_twins)

# Now we can use p_first_twin as a prior for the second birth twin
p_second_twin <- p_first_twin * p_twins

# probability of the next birth is twin when we don't know what specie we have
sum(p_second_twin)

```

2H2. Recall all the facts from the problem above. Now compute the probability that the panda we
have is from species A, assuming we have observed only the first birth and that it was twins.

```{r}
# Define A and B species probability of getting twins
p_twins <- c(0.1, 0.2)

priors <- c(1, 1)

posterior <- (p_twins * priors) / sum(p_twins)

posterior[1]
```

```{r}
# Or using Byes Theorem 

p_A <- 0.5

p_tA <- 0.1
p_tB <- 0.2

# average probability of twins
p_t <- (p_A * p_tA) + ((1 - p_A) * p_tB)

# P(A|twins)
p_tA * p_A / p_t
```


2H3. Continuing on from the previous problem, suppose the same panda mother has a second birth
and that it is not twins, but a singleton infant. Compute the posterior probability that this panda is species A.

```{r}
p_twins <- c(0.1, 0.2)
p_single <- 1 - p_twins

priors <- c(1, 1)

p_first_twin <- (p_twins * priors) / sum(p_twins)

p_second_single <- (p_first_twin * p_single)
p_second_single <- posterior / sum(posterior)
p_second_single[1]
```

  2H4. A common boast of Bayesian statisticians is that Bayesian inference makes it easy to use all of the data, even if the data are of different types.
  So suppose now that a veterinarian comes along who has a new genetic test that she claims can
identify the species of our mother panda. But the test, like all tests, is imperfect. This is the information you have about the test:
• The probability it correctly identifies a species A panda is 0.8.
• The probability it correctly identifies a species B panda is 0.65.
  The vet administers the test to your panda and tells you that the test is positive for species A. First ignore your previous information from the births and compute the posterior probability that your panda is species A. Then redo your calculation, now using the birth data as well.


```{r}
# First withour any birth data
prior <- c(1, 1)

# probability of A given test 
p_a <- c(0.8, 0.35)

test_posterior <- p_a * prior / sum(p_a)

# Prob to have A if test says A
test_posterior[1]
```

```{r}
# include birth data 

p_twins <- c(0.1, 0.2)
p_single <- 1 - p_twins

# use posterior from the last chunk as a prior 
test_prior <- test_posterior

posterior <- test_prior * p_twins * p_single
posterior <- posterior / sum(posterior)
posterior[1]
```


Additional task from 
https://github.com/rmcelreath/statrethinking_winter2019/blob/master/homework/week01.pdf

3. This problem is more open-ended than the others. Feel free to collaborate on the solution. Suppose you want to estimate the Earths proportion of water very precisely. Specifically, you want the 99% percentile interval of the posterior distribution to be only 0.05 wide. This means the distance between the upper and lower bound of the interval should be 0.05. How many times will you have to toss the globe to do this? I won’t require a precise answer. I’m honestly more interested in your approach.



```{r}
# define the true water proportion
water_prop <- 0.71

# define the vector in which number of tosses needed will be stored
minimum_n_tosses <- rep(NA, 100)


for (i in 1:100) {
  
  # start from 100 tosses
  n_toss <- 100
  desired_pi_width <- 0.05
  actual_pi_width <- 1

  
  while (actual_pi_width > desired_pi_width) {
    
    # randomly generate globe tosses
    tosses <- rbinom(1e4, size = n_toss, prob = water_prop)
    
    prop_earth <- tosses / n_toss
    
    # calculate actual 99% interval
    upper_bound <- quantile(prop_earth, probs = c(0.995))
    lower_bound <- quantile(prop_earth, probs = c(0.005))
    
    actual_pi_width <- upper_bound - lower_bound
    
    if (actual_pi_width < desired_pi_width) {
      minimum_n_tosses[i] <- n_toss
      break
    }
    
    n_toss <- n_toss + 10
  }
}

```

```{r}
boxplot(minimum_n_tosses)
```

It seems that to have 99% percentile interval of the posterior distribution to be
only 0.05 wide we will have to toss the globe more then 2000 times. 
Assuming that each toss is independent from other. 
