---
title: "2nd Chapter"
author: "Mykola Dereva"
date: "1/11/2021"
output: html_document
---

```{r}
library(ggplot2)
library(tidyr)
library(ggeasy)
library(dplyr)
```



2E1. Which of the expressions below correspond to the statement: the probability of rain on Monday?

  (1) Pr(rain)
  (2) Pr(rain|Monday)
  (3) Pr(Monday|rain)
  (4) Pr(rain, Monday)/ Pr(Monday)

Answer: 2, 4

2E2. Which of the following statements corresponds to the expression: Pr(Monday|rain)?

  (1) The probability of rain on Monday.
  (2) The probability of rain, given that it is Monday.
  (3) The probability that it is Monday, given that it is raining.
  (4) The probability that it is Monday and that it is raining.

Answer: 3

2E3. Which of the expressions below correspond to the statement: the probability that it is Monday,
given that it is raining?

  (1) Pr(Monday|rain)
  (2) Pr(rain|Monday)
  (3) Pr(rain|Monday) Pr(Monday)
  (4) Pr(rain|Monday) Pr(Monday)/ Pr(rain)
  (5) Pr(Monday|rain) Pr(rain)/ Pr(Monday)

Answer: 1

2E4. The Bayesian statistician Bruno de Finetti (1906–1985) began his 1973 book on probability theory with the declaration: “PROBABILITY DOES NOT EXIST.” The capitals appeared in the original,
so I imagine de Finetti wanted us to shout this statement. What he meant is that probability is a device for describing uncertainty from the perspective of an observer with limited knowledge; it has no objective reality. Discuss the globe tossing example from the chapter, in light of this statement. What does it mean to say “the probability of water is 0.7”?

Answer: As for me, this means that even though we do not know the "true" proportion of the globe
covered with the water, but given the observations we have, our best guess is that there is a 70% chance of getting a "water" in the next toss of a globe.
In other words, we can estimate the continuous range of probabilities even if in each toss we
can get only a binary outcome (0 - land; 1 - water). 
This probability helps us to understand the proportion of a water on the Earth, however
it exists only in our imagination. 


2M1. Recall the globe tossing model from the chapter. Compute and plot the grid approximate
posterior distribution for each of the following sets of observations. In each case, assume a uniform prior for p.
  (1) W, W, W
  (2) W, W, W, L
  (3) L, W, W, L, W, W, W


```{r}
# define grid length 
grid_l <- 30

# define grid
p_grid <- seq(from = 0 , to = 1 , length.out = grid_l)

# define prior
prior <- rep(1, grid_l)

# compute likelihood at each value in grid
likelihood_1 <- dbinom(3, size = 3, prob = p_grid)
likelihood_2 <- dbinom(3, size = 4, prob = p_grid)
likelihood_3 <- dbinom(5, size = 7, prob = p_grid)

# compute product of likelihood and prior
unstd.posterior_1 <- likelihood_1 * prior
unstd.posterior_2 <- likelihood_2 * prior
unstd.posterior_3 <- likelihood_3 * prior

# standardize the posterior, so it sums to 1
posterior_1 <- unstd.posterior_1 / sum(unstd.posterior_1)
posterior_2 <- unstd.posterior_2 / sum(unstd.posterior_2)
posterior_3 <- unstd.posterior_3 / sum(unstd.posterior_3)
```

```{r}
# Plot a results
data.frame(p_grid, posterior_1, posterior_2, posterior_3) %>%
  pivot_longer(cols = starts_with("post"), names_to = "posterior", values_to = "prob") %>%
  mutate(posterior = case_when(posterior == "posterior_1" ~ "W, W, W",
                               posterior == "posterior_2" ~ "W, W, W, L",
                               posterior == "posterior_3" ~ "L, W, W, L, W, W, W")) %>% 
  ggplot(aes(x = p_grid, y = prob, col = posterior)) +
  geom_point() +
  geom_line() +
  labs(x = "Proportion of Water",
       y = "Posterior Probability",
       color = "Pattern") +
  theme_minimal()


```



2M2. Now assume a prior for p that is equal to zero when p < 0.5 and is a positive constant when
p ≥ 0.5. Again compute and plot the grid approximate posterior distribution for each of the sets of
observations in the problem just above