---
title: "3rd Chapter"
author: "Mykola Dereva"
date: "1/14/2021"
output: html_document
---

```{r}
p_grid <- seq(from = 0 , to = 1 , length.out = 1000)
prior <- rep( 1 , 1000 )
likelihood <- dbinom(6 , size = 9 , prob = p_grid)
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)
set.seed(100)
samples <- sample(p_grid , prob = posterior , size = 1e4, replace = TRUE)
```

```{r}
plot(density(samples))
```

Use the values in samples to answer the questions that follow

3E1. How much posterior probability lies below p = 0.2?
```{r}
mean(samples < 0.5) 
```

3E2. How much posterior probability lies above p = 0.8?
```{r}
mean(samples > 0.8) 
```

3E3. How much posterior probability lies between p = 0.2 and p = 0.8?

```{r}
mean(samples > 0.2 & samples < 0.8)
```


3E4. 20% of the posterior probability lies below which value of p?
```{r}
quantile(samples, prob = c(0.2))
```


3E5. 20% of the posterior probability lies above which value of p?
```{r}
quantile(samples, prob = c(0.8))
```

3E6. Which values of p contain the narrowest interval equal to 66% of the posterior probability?
```{r}
rethinking::HPDI(samples, prob = 0.66)
```

3E7. Which values of p contain 66% of the posterior probability, assuming equal posterior probability both below and above the interval?

```{r}
rethinking::PI(samples, prob = 0.66)
```

3M1. Suppose the globe tossing data had turned out to be 8 water in 15 tosses. Construct the posterior distribution, using grid approximation. Use the same flat prior as before.

```{r}
p_grid <- seq(from = 0 , to = 1 , length.out = 1000)
prior <- rep(1 , 1000)
likelihood <- dbinom(8 , size = 15, prob = p_grid)
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)
```


3M2. Draw 10,000 samples from the grid approximation from above. Then use the samples to calculate the 90% HPDI for p

```{r}
set.seed(100)
samples <- sample(p_grid , prob = posterior , size = 1e4, replace = TRUE)

rethinking::HPDI(samples, prob = 0.9)
```

3M3. Construct a posterior predictive check for this model and data. This means simulate the distribution of samples, averaging over the posterior uncertainty in p. What is the probability of observing 8 water in 15 tosses?