---
title: "5th Chapter BRMS"
author: "Mykola Dereva"
date: "3/29/2021"
output: html_document
---


```{r}
rm(list = ls())
library(tidyverse)

data(WaffleDivorce, package = "rethinking")
d <- WaffleDivorce
```



```{r}
d <-
  d %>% 
  mutate(d = rethinking::standardize(Divorce),
         m = rethinking::standardize(Marriage),
         a = rethinking::standardize(MedianAgeMarriage))
```


```{r}
rm(WaffleDivorce)
library(brms)
library(ggrepel)  # text lables on plot 

```

```{r}
glimpse(d)
```


```{r}
d %>%
  ggplot(aes(x = WaffleHouses/Population, y = Divorce)) +
  stat_smooth(method = "lm", fullrange = T, size = 1/2,
              color = "firebrick4", fill = "firebrick", alpha = 1/5) +
  geom_point(size = 3, color = "firebrick4", alpha = 1/2, shape = 1) +
  geom_text_repel(data = d %>% filter(Loc %in% #specify labels
                                        c("ME", "OK", "AR", "AL", "GA", "SC", "NJ")),  
                  aes(label = Loc), 
                  size = 3, seed = 1042) +  # this makes it reproducible
  scale_x_continuous("Waffle Houses per million", limits = c(0, 55)) +
  ylab("Divorce rate") +
  coord_cartesian(xlim = c(0, 50), ylim = c(5, 15)) +
  theme_bw() +
  theme(panel.grid = element_blank())
```



```{r}
sd(d$MedianAgeMarriage)
```


```{r}
b5.1 <- 
  brm(data = d, 
      family = gaussian,
      d ~ 1 + a,
      prior = c(prior(normal(0, 0.2), class = Intercept),
                prior(normal(0, 0.5), class = b),
                prior(exponential(1), class = sigma)),
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      seed = 5,
      sample_prior = T,     # automate prior predictive simulation
      file = "fits/b05.01")
```


```{r}
prior <- prior_samples(b5.1)    # now we have prior samples 

prior %>% glimpse()
```


```{r}
set.seed(5)

prior %>% 
  slice_sample(n = 200) %>%         # number of prior samples
  rownames_to_column("draw") %>%       # id of each line
  expand(nesting(draw, Intercept, b),
         a = c(-2, 2)) %>% 
  mutate(d = Intercept + b * a) %>% 
  
  ggplot(aes(x = a, y = d)) +
  geom_line(aes(group = draw),
            color = "firebrick", alpha = 0.1) +
  labs(x = "Median age marriage (std)",
       y = "Divorce rate (std)") +
  ggtitle("Prior predictive simulation") +
  coord_cartesian(ylim = c(-2, 2)) +
  theme_bw() +
  theme(panel.grid = element_blank()) 
```

Posterior predictive simulation

```{r}
# determine the range of `a` values we'd like to feed into `fitted()`
nd <- tibble(a = seq(from = -3, to = 3.2, length.out = 30))

# now use `fitted()` to get the model-implied trajectories
fitted(b5.1,
       newdata = nd) %>% 
  data.frame() %>% 
  bind_cols(nd) %>% 
  
  # plot
  ggplot(aes(x = a)) +
  geom_smooth(aes(y = Estimate, ymin = Q2.5, ymax = Q97.5),
              stat = "identity",
              fill = "firebrick", color = "firebrick4", alpha = 1/5, size = 1/4) +
  geom_point(data = d, 
             aes(y = d), 
             size = 3, color = "firebrick4", shape = 1) +
  labs(x = "Median age marriage (std)",
       y = "Divorce rate (std)") +
  ggtitle("Posterior predictive simulation") +
  coord_cartesian(xlim = range(d$a), 
                  ylim = range(d$d)) +
  theme_bw() +
  theme(panel.grid = element_blank()) 
```


```{r}
print(b5.1)
```



```{r}
b5.2 <- 
  brm(data = d, 
      family = gaussian,
      d ~ 1 + m,
      prior = c(prior(normal(0, 0.2), class = Intercept),
                prior(normal(0, 0.5), class = b),
                prior(exponential(1), class = sigma)),
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      seed = 5,
      file = "fits/b05.02")
```


```{r}
print(b5.2)
```



```{r}
nd <- tibble(m = seq(from = -2.5, to = 3.5, length.out = 30))

fitted(b5.2, newdata = nd) %>%
  data.frame() %>%
  bind_cols(nd) %>% 
  
  ggplot(aes(x = m)) +
  geom_smooth(aes(y = Estimate, ymin = Q2.5, ymax = Q97.5),
              stat = "identity",
              fill = "firebrick", color = "firebrick4", alpha = 1/5, size = 1/4) +
  geom_point(data = d, 
             aes(y = d), 
             size = 3, color = "firebrick4", shape = 1) +
  labs(x = "Marriage rate (std)",
       y = "Divorce rate (std)") +
  coord_cartesian(xlim = range(d$m), 
                  ylim = range(d$d)) +
  theme_bw() +
  theme(panel.grid = element_blank())                   
```




## DAGs

```{r}
library(ggdag)
set.seed(5)

dagify(M ~ A,
       D ~ A + M) %>%
  ggdag(node_size = 8)
```



```{r}
dag_coords <-
  tibble(name = c("A", "M", "D"),
         x    = c(1, 3, 2),
         y    = c(2, 2, 1))

p1 <-
  dagify(M ~ A,
         D ~ A + M,
         coords = dag_coords) %>%
  
  ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_point(color = "firebrick", alpha = 1/4, size = 10) +
  geom_dag_text(color = "firebrick") +
  geom_dag_edges(edge_color = "firebrick") +
  scale_x_continuous(NULL, breaks = NULL, expand = c(.1, .1)) +
  scale_y_continuous(NULL, breaks = NULL, expand = c(.1, .1)) +
  theme_bw() +
  theme(panel.grid = element_blank())

p1
```


```{r}
p2 <-
  dagify(M ~ A,
         D ~ A,
         coords = dag_coords) %>%
  
  ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_point(color = "firebrick", alpha = 1/4, size = 10) +
  geom_dag_text(color = "firebrick") +
  geom_dag_edges(edge_color = "firebrick") +
  scale_x_continuous(NULL, breaks = NULL, expand = c(.1, .1)) +
  scale_y_continuous(NULL, breaks = NULL, expand = c(.1, .1)) +
  theme_bw() +
  theme(panel.grid = element_blank()) 

p2
```


```{r}
library(patchwork)

p1 | p2
```



```{r}
library(dagitty)

dagitty('dag{ D <- A -> M }') %>% 
  impliedConditionalIndependencies()
```

```{r}
dagitty('dag{D <- A -> M -> D}') %>% 
  impliedConditionalIndependencies()
```




## Multiple regression 


```{r}
b5.3 <- 
  brm(data = d, 
      family = gaussian,
      d ~ 1 + m + a,
      prior = c(prior(normal(0, 0.2), class = Intercept),
                prior(normal(0, 0.5), class = b),
                prior(exponential(1), class = sigma)),
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      seed = 5,
      file = "fits/b05.03")
```

```{r}
print(b5.3)
```



```{r}
# first, extract and rename the necessary posterior parameters
bind_cols(
  posterior_samples(b5.1) %>% 
    transmute(`b5.1_beta[A]` = b_a),
  posterior_samples(b5.2) %>% 
    transmute(`b5.2_beta[M]` = b_m),
  posterior_samples(b5.3) %>% 
    transmute(`b5.3_beta[M]` = b_m,
              `b5.3_beta[A]` = b_a)
  ) %>% 
  # convert them to the long format, group, and get the posterior summaries
  pivot_longer(everything()) %>% 
  group_by(name) %>% 
  summarise(mean = mean(value),
            ll   = quantile(value, prob = .025),
            ul   = quantile(value, prob = .975)) %>% 
  # since the `key` variable is really two variables in one, here we split them up
  separate(col = name, into = c("fit", "parameter"), sep = "_") %>% 
  
  # plot!
  ggplot(aes(x = mean, xmin = ll, xmax = ul, y = fit)) +
  geom_vline(xintercept = 0, color = "firebrick", alpha = 1/5) +
  geom_pointrange(color = "firebrick") +
  labs(x = "posterior", y = NULL) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        strip.background = element_rect(fill = "transparent", color = "transparent")) +
  facet_wrap(~ parameter, ncol = 1, labeller = label_parsed)
```




##  Predictor residual plots


```{r}
b5.4 <- 
  brm(data = d, 
      family = gaussian,
      m ~ 1 + a,
      prior = c(prior(normal(0, 0.2), class = Intercept),
                prior(normal(0, 0.5), class = b),
                prior(exponential(1), class = sigma)),
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      seed = 5,
      file = "fits/b05.04")
```



```{r}
print(b5.4)
```



```{r}
f <- 
  fitted(b5.4) %>%
  data.frame() %>%
  bind_cols(d)

glimpse(f)
```



```{r}
p1 <-
  f %>% 
  ggplot(aes(x = a, y = m)) +
  geom_point(size = 3, shape = 1, color = "firebrick4") +
  geom_segment(aes(xend = a, yend = Estimate), 
               size = 1/4) +
  geom_line(aes(y = Estimate), 
            color = "firebrick4") +
  geom_text_repel(data = . %>% filter(Loc %in% c("WY", "ND", "ME", "HI", "DC")),  
                  aes(label = Loc), 
                  size = 3, seed = 14) +
  labs(x = "Age at marriage (std)",
       y = "Marriage rate (std)") +
  coord_cartesian(ylim = range(d$m)) +
  theme_bw() +
  theme(panel.grid = element_blank()) 

p1
```



```{r}
r <- 
  residuals(b5.4) %>%
  # to use this in ggplot2, we need to make it a tibble or data frame
  data.frame() %>% 
  bind_cols(d)

p3 <-
  r %>% 
  ggplot(aes(x = Estimate, y = d)) +
  stat_smooth(method = "lm", fullrange = T,
              color = "firebrick4", fill = "firebrick4", 
              alpha = 1/5, size = 1/2) +
  geom_vline(xintercept = 0, linetype = 2, color = "grey50") +
  geom_point(size = 3, color = "firebrick4", alpha = 2/3, shape = 1) +
  geom_text_repel(data = . %>% filter(Loc %in% c("WY", "ND", "ME", "HI", "DC")),  
                  aes(label = Loc), 
                  size = 3, seed = 5) +
  scale_x_continuous(limits = c(-2, 2)) +
  coord_cartesian(xlim = range(r$Estimate)) +
  labs(x = "Marriage rate residuals",
       y = "Divorce rate (std)") +
  theme_bw() +
  theme(panel.grid = element_blank())

p3
```



```{r}
b5.4b <- 
  brm(data = d, 
      family = gaussian,
      a ~ 1 + m,
      prior = c(prior(normal(0, 0.2), class = Intercept),
                prior(normal(0, 0.5), class = b),
                prior(exponential(1), class = sigma)),
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      seed = 5,
      file = "fits/b05.04b")
```



```{r}
p2 <-
  fitted(b5.4b) %>%
  data.frame() %>%
  bind_cols(d) %>% 
  
  ggplot(aes(x = m, y = a)) +
  geom_point(size = 3, shape = 1, color = "firebrick4") +
  geom_segment(aes(xend = m, yend = Estimate), 
               size = 1/4) +
  geom_line(aes(y = Estimate), 
            color = "firebrick4") +
  geom_text_repel(data = . %>% filter(Loc %in% c("DC", "HI", "ID")),  
                  aes(label = Loc), 
                  size = 3, seed = 5) +
  labs(x = "Marriage rate (std)",
       y = "Age at marriage (std)") +
  coord_cartesian(ylim = range(d$a)) +
  theme_bw() +
  theme(panel.grid = element_blank())   

p2
```


```{r}
r <-
  residuals(b5.4b) %>%
  data.frame() %>% 
  bind_cols(d)

p4 <-
  r %>%
  ggplot(aes(x = Estimate, y = d)) +
  stat_smooth(method = "lm", fullrange = T,
              color = "firebrick4", fill = "firebrick4", 
              alpha = 1/5, size = 1/2) +
  geom_vline(xintercept = 0, linetype = 2, color = "grey50") +
  geom_point(size = 3, color = "firebrick4", alpha = 2/3, shape = 1) +
  geom_text_repel(data = . %>% filter(Loc %in% c("ID", "HI", "DC")),  
                  aes(label = Loc), 
                  size = 3, seed = 5) +
  scale_x_continuous(limits = c(-2, 3)) +
  coord_cartesian(xlim = range(r$Estimate),
                  ylim = range(d$d)) +
  labs(x = "Age at marriage residuals",
       y = "Divorce rate (std)") +
  theme_bw() +
  theme(panel.grid = element_blank())

p4
```



```{r}
p1 + p2 + p3 + p4 + plot_annotation(title = "Understanding multiple regression through residuals")
```


## Posterior prediction plot

```{r}
fitted(b5.3) %>%
  data.frame() %>%
  # un-standardize the model predictions
  mutate_all(~. * sd(d$Divorce) + mean(d$Divorce)) %>% 
  bind_cols(d) %>%
  
  ggplot(aes(x = Divorce, y = Estimate)) +
  geom_abline(linetype = 2, color = "grey50", size = .5) +
  geom_point(size = 3, color = "firebrick4", alpha = 3/4) +
  geom_linerange(aes(ymin = Q2.5, ymax = Q97.5),
                 size = 1/4, color = "firebrick4") +
  geom_text(data = . %>% filter(Loc %in% c("ID", "UT", "RI", "ME")),
            aes(label = Loc), 
            hjust = 1, nudge_x = - 0.25) +
  labs(x = "Observed divorce", y = "Predicted divorce") +
  theme_bw() +
  theme(panel.grid = element_blank())
```

## Counterfactual plots

Here we have to use multivariate plot in brms 
more on it here: 
https://cran.r-project.org/package=brms/vignettes/brms_multivariate.html


First specify two models 
```{r}
d_model <- bf(d ~ 1 + a + m)
m_model <- bf(m ~ 1 + a)
```


Second, include both of them into one model
```{r}
b5.3_A <-
  brm(data = d, 
      family = gaussian,
      d_model + m_model + set_rescor(FALSE),
      prior = c(prior(normal(0, 0.2), class = Intercept, resp = d),
                prior(normal(0, 0.5), class = b,         resp = d),
                prior(exponential(1), class = sigma,     resp = d),
                
                prior(normal(0, 0.2), class = Intercept, resp = m),
                prior(normal(0, 0.5), class = b,         resp = m),
                prior(exponential(1), class = sigma,     resp = m)),
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      seed = 5,
      file = "fits/b05.03_A")
```



```{r}
print(b5.3_A)
```



```{r}
nd <- tibble(a = seq(from = -2, to = 2, length.out = 30),  # "a" vary
             m = 0)                                        # "m" constant

p1 <-
  predict(b5.3_A,
          resp = "d",
          newdata = nd) %>% 
  data.frame() %>% 
  bind_cols(nd) %>% 
  
  ggplot(aes(x = a, y = Estimate, ymin = Q2.5, ymax = Q97.5)) +
  geom_smooth(stat = "identity",
              fill = "firebrick", color = "firebrick4", alpha = 1/5, size = 1/4) +
  labs(subtitle = "Total counterfactual effect of A on D",
       x = "manipulated A",
       y = "counterfactual D") +
  coord_cartesian(ylim = c(-2, 2)) +
  theme_bw() +
  theme(panel.grid = element_blank()) 


```

```{r}
nd <- tibble(a = seq(from = -2, to = 2, length.out = 30))

p2 <-
  predict(b5.3_A,
          resp = "m",
          newdata = nd) %>% 
  data.frame() %>% 
  bind_cols(nd) %>% 
  
  ggplot(aes(x = a, y = Estimate, ymin = Q2.5, ymax = Q97.5)) +
  geom_smooth(stat = "identity",
              fill = "firebrick", color = "firebrick4", alpha = 1/5, size = 1/4) +
  labs(subtitle = "Counterfactual effect of A on M",
       x = "manipulated A",
       y = "counterfactual M") +
  coord_cartesian(ylim = c(-2, 2)) +
  theme_bw() +
  theme(panel.grid = element_blank()) 

p1 + p2 + plot_annotation(title = "Counterfactual plots for the multivariate divorce model")
```



```{r}
nd <- tibble(m = seq(from = -2, to = 2, length.out = 30),
             a = 0)

predict(b5.3_A,
        resp = "d",
        newdata = nd) %>% 
  data.frame() %>% 
  bind_cols(nd) %>% 
  
  ggplot(aes(x = m, y = Estimate, ymin = Q2.5, ymax = Q97.5)) +
  geom_smooth(stat = "identity",
              fill = "firebrick", color = "firebrick4", alpha = 1/5, size = 1/4) +
  labs(subtitle = "Total counterfactual effect of M on D",
       x = "manipulated M",
       y = "counterfactual D") +
  coord_cartesian(ylim = c(-2, 2)) +
  theme_bw() +
  theme(panel.grid = element_blank()) 
```




```{r}
data(milk, package = "rethinking")
d <- milk
rm(milk)

glimpse(d)
```



```{r}
d <-
  d %>% 
  mutate(kcal.per.g_s     = (kcal.per.g - mean(kcal.per.g)) / sd(kcal.per.g), 
         log_mass_s       = (log(mass) - mean(log(mass))) / sd(log(mass)), 
         neocortex.perc_s = (neocortex.perc - mean(neocortex.perc, na.rm = T)) / sd(neocortex.perc, na.rm = T))
```



```{r}
dcc <- 
  d %>%
  drop_na(ends_with("_s"))

# how many rows did we drop?
nrow(d) - nrow(dcc)
```



```{r}
b5.5_draft <- 
  brm(data = dcc, 
      family = gaussian,
      kcal.per.g_s ~ 1 + neocortex.perc_s,
      prior = c(prior(normal(0, 1), class = Intercept),
                prior(normal(0, 1), class = b),
                prior(exponential(1), class = sigma)),
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      seed = 5,
      sample_prior = T,
      file = "fits/b05.05_draft")
```



```{r}
set.seed(5)

prior_samples(b5.5_draft) %>% 
  slice_sample(n = 50) %>% 
  rownames_to_column() %>% 
  expand(nesting(rowname, Intercept, b),
         neocortex.perc_s = c(-2, 2)) %>% 
  mutate(kcal.per.g_s = Intercept + b * neocortex.perc_s) %>% 
  
  ggplot(aes(x = neocortex.perc_s, y = kcal.per.g_s)) +
  geom_line(aes(group = rowname),
            color = "firebrick", alpha = .4) +
  coord_cartesian(ylim = c(-2, 2)) +
  labs(x = "neocortex percent (std)",
       y = "kilocal per g (std)",
       subtitle = "Intercept ~ dnorm(0, 1)\nb ~ dnorm(0, 1)") +
  theme_bw() +
  theme(panel.grid = element_blank()) 
```



```{r}
b5.5 <- 
  brm(data = dcc, 
      family = gaussian,
      kcal.per.g_s ~ 1 + neocortex.perc_s,
      prior = c(prior(normal(0, 0.2), class = Intercept),
                prior(normal(0, 0.5), class = b),
                prior(exponential(1), class = sigma)),
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      seed = 5,
      sample_prior = T,
      file = "fits/b05.05")
```

```{r}
set.seed(5)

prior_samples(b5.5) %>% 
  slice_sample(n = 50) %>% 
  rownames_to_column() %>% 
  expand(nesting(rowname, Intercept, b),
         neocortex.perc_s = c(-2, 2)) %>% 
  mutate(kcal.per.g_s = Intercept + b * neocortex.perc_s) %>% 
  
  ggplot(aes(x = neocortex.perc_s, y = kcal.per.g_s)) +
  geom_line(aes(group = rowname),
            color = "firebrick", alpha = .4) +
  coord_cartesian(ylim = c(-2, 2)) +
  labs(x = "neocortex percent (std)",
       y = "kilocal per g (std)",
       subtitle = "Intercept ~ dnorm(0, 0.2)\nb ~ dnorm(0, 0.5)") +
  theme_bw() +
  theme(panel.grid = element_blank()) 
```



```{r}
b5.6 <- 
  brm(data = dcc, 
      family = gaussian,
      kcal.per.g_s ~ 1 + log_mass_s,
      prior = c(prior(normal(0, 0.2), class = Intercept),
                prior(normal(0, 0.5), class = b),
                prior(exponential(1), class = sigma)),
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      seed = 5,
      sample_prior = T,
      file = "fits/b05.06")
```


```{r}
print(b5.6)
```


```{r}
nd <- tibble(log_mass_s = seq(from = -2.5, to = 2.5, length.out = 30))

fitted(b5.6, 
       newdata = nd,
       probs = c(.025, .975, .25, .75)) %>%
  data.frame() %>%
  bind_cols(nd) %>% 
  
  ggplot(aes(x = log_mass_s, y = Estimate)) +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5),
              fill = "firebrick", alpha = 1/5) +
  geom_smooth(aes(ymin = Q25, ymax = Q75),
              stat = "identity",
              fill = "firebrick4", color = "firebrick4", alpha = 1/5, size = 1/2) +
  geom_point(data = dcc, 
             aes(y = kcal.per.g_s),
             size = 2, color = "firebrick4") +
  coord_cartesian(xlim = range(dcc$log_mass_s), 
                  ylim = range(dcc$kcal.per.g_s)) +
  labs(x = "log body mass (std)",
       y = "kilocal per g (std)") +
  theme_bw() +
  theme(panel.grid = element_blank())
```



```{r}
b5.7 <- 
  brm(data = dcc, 
      family = gaussian,
      kcal.per.g_s ~ 1 + neocortex.perc_s + log_mass_s,
      prior = c(prior(normal(0, 0.2), class = Intercept),
                prior(normal(0, 0.5), class = b),
                prior(exponential(1), class = sigma)),
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      seed = 5,
      file = "fits/b05.07")
```


```{r}
print(b5.7)
```



```{r}
library(GGally)

dcc %>% 
  select(ends_with("_s")) %>% 
  ggpairs()
```



Conterfactual plots 

```{r}
nd <- tibble(neocortex.perc_s = seq(from = -2.5, to = 2, length.out = 30),
             log_mass_s       = 0)

p1 <-
  fitted(b5.7, 
         newdata = nd,
         probs = c(.025, .975, .25, .75)) %>%
  data.frame() %>%
  bind_cols(nd) %>% 
  
  ggplot(aes(x = neocortex.perc_s, y = Estimate)) +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5),
              fill = "firebrick", alpha = 1/5) +
  geom_smooth(aes(ymin = Q25, ymax = Q75),
              stat = "identity",
              fill = "firebrick4", color = "firebrick4", alpha = 1/5, size = 1/2) +
  coord_cartesian(xlim = range(dcc$neocortex.perc_s), 
                  ylim = range(dcc$kcal.per.g_s)) +
  labs(subtitle = "Counterfactual holding M = 0", 
       x = "neocortex percent (std)",
       y = "kilocal per g (std)")
```


```{r}
nd <- tibble(log_mass_s       = seq(from = -2.5, to = 2.5, length.out = 30),
             neocortex.perc_s = 0)

p2 <-
  fitted(b5.7, 
         newdata = nd,
         probs = c(.025, .975, .25, .75)) %>%
  data.frame() %>%
  bind_cols(nd) %>% 
  
  ggplot(aes(x = log_mass_s, y = Estimate)) +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5),
              fill = "firebrick", alpha = 1/5) +
  geom_smooth(aes(ymin = Q25, ymax = Q75),
              stat = "identity",
              fill = "firebrick4", color = "firebrick4", alpha = 1/5, size = 1/2) +
  coord_cartesian(xlim = range(dcc$log_mass_s), 
                  ylim = range(dcc$kcal.per.g_s)) +
  labs(subtitle = "Counterfactual holding N = 0",
       x = "log body mass (std)",
       y = "kilocal per g (std)")

```


```{r}
# combine
p1 + p2 + 
  plot_annotation(title = "Figure 5.9 [bottom row]. Milk energy and neocortex among primates.") &
  theme_bw() &
  theme(panel.grid = element_blank())
```



## Catigorical variables 

```{r}
data(Howell1, package = "rethinking")
d <- Howell1
rm(Howell1)
```

```{r}
d %>% glimpse()
```


```{r}
set.seed(5)

prior <-
  tibble(mu_female = rnorm(1e4, mean = 178, sd = 20)) %>% 
  mutate(mu_male = mu_female + rnorm(1e4, mean = 0, sd = 10))

prior %>% 
  pivot_longer(everything()) %>% 
  group_by(name) %>% 
  summarise(mean = mean(value),
            sd   = sd(value),
            ll   = quantile(value, prob = .025),
            ul   = quantile(value, prob = .975)) %>% 
  mutate_if(is.double, round, digits = 2)
```


```{r}
prior %>% 
  pivot_longer(everything()) %>% 
  ggplot(aes(x = value, fill = name, color = name)) +
  geom_density(size = 2/3, alpha = 0.25) +
  scale_fill_manual(NULL, values = c("firebrick4", "black")) +
  scale_color_manual(NULL, values = c("firebrick4", "black")) +
  scale_y_continuous(NULL, breaks = NULL) +
  xlab("prior predictive distribution for our dummy groups") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        legend.position = c(.82, .83))
```


```{r}
d <-
  d %>% 
  mutate(sex = ifelse(male == 1, 2, 1))

head(d)
```


```{r}
d <-
  d %>% 
  mutate(sex = factor(sex))
```


```{r}
b5.8 <- 
  brm(data = d, 
      family = gaussian,
      height ~ 0 + sex,   # 0 means that we have separate intercept for each sex
      prior = c(prior(normal(178, 20), class = b),
                prior(exponential(1), class = sigma)),
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      seed = 5,
      file = "fits/b05.08")
```
```{r}
print(b5.8)
```


```{r}
library(tidybayes)

posterior_samples(b5.8) %>% 
  mutate(diff_fm = b_sex1 - b_sex2) %>% 
  pivot_longer(everything()) %>% 
  filter(name != "lp__") %>% 
  group_by(name) %>% 
  mean_qi(value, .width = .95)
```



## Multiple categories 


```{r}
data(milk, package = "rethinking")
d <- milk
rm(milk)
```

```{r}
d %>% glimpse()
```

```{r}
d %>% distinct(clade)
```



```{r}
d <-
  d %>% 
  mutate(kcal.per.g_s = (kcal.per.g - mean(kcal.per.g)) / sd(kcal.per.g))
```



```{r}
b5.9 <- 
  brm(data = d, 
      family = gaussian,
      kcal.per.g_s ~ 0 + clade,
      prior = c(prior(normal(0, 0.5), class = b),
                prior(exponential(1), class = sigma)),
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      seed = 5,
      file = "fits/b05.09")
```


```{r}
print(b5.9)
```


```{r}
mcmc_plot(b5.9, pars = "^b_")
```



```{r}
library(bayesplot)

color_scheme_set("red")

post <- posterior_samples(b5.9)

post %>% 
  select(starts_with("b_")) %>% 
  mcmc_intervals(prob = .5,
                 point_est = "median") +
  labs(title = "My fancy bayesplot-based coefficient plot") +
  theme_bw() +
  theme(axis.text.y = element_text(hjust = 0),
        axis.ticks.y = element_blank(),
        panel.grid = element_blank())
```



```{r}
library(tidybayes)

post %>% 
  select(starts_with("b")) %>% 
  set_names(distinct(d, clade) %>% arrange(clade) %>% pull()) %>% 
  pivot_longer(everything()) %>% 
  
  ggplot(aes(x = value, y = reorder(name, value))) +  # note how we used `reorder()` to arrange the coefficients
  geom_vline(xintercept = 0, color = "firebrick4", alpha = 1/10) +
  stat_pointinterval(point_interval = mode_hdi, .width = .89, 
                     size = 1, color = "firebrick4") +
  labs(title = "My tidybayes-based coefficient plot",
       x = "expected kcal (std)", 
       y = NULL) +
  theme_bw() +
  theme(axis.text.y = element_text(hjust = 0),
        axis.ticks.y = element_blank(),
        panel.grid = element_blank())
```



```{r}
houses <- c("Gryffindor", "Hufflepuff", "Ravenclaw", "Slytherin")

set.seed(63)
d <-
  d %>% 
  mutate(house = sample(rep(houses, each = 8), size = n()))
```



```{r}
b5.10 <- 
  brm(data = d, 
      family = gaussian,
      kcal.per.g_s ~ 0 + clade + house,
      prior = c(prior(normal(0, 0.5), class = b),
                prior(exponential(1), class = sigma)),
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      seed = 5,
      file = "fits/b05.10")
```


```{r}
print(b5.10)
```

Here one of the houses got dropped 
for details check: https://bookdown.org/content/4857/the-many-variables-the-spurious-waffles.html#categorical-variables 

To have intercept for each variable we have to use nonlinear brms syntax 

```{r}
b5.11 <- 
  brm(data = d, 
      family = gaussian,
      bf(kcal.per.g_s ~ 0 + a + h, 
         a ~ 0 + clade,                # here 0 for each group of categorical
         h ~ 0 + house,
         nl = TRUE),
      prior = c(prior(normal(0, 0.5), nlpar = a),
                prior(normal(0, 0.5), nlpar = h),
                prior(exponential(1), class = sigma)),
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      seed = 5,
      file = "fits/b05.11")
```

```{r}
print(b5.11)
```

Here we have 4 coefs for clade and 4 for house, as we expected 


```{r}
posterior_samples(b5.11) %>% 
  pivot_longer(starts_with("b_")) %>% 
  mutate(name = str_remove(name, "b_") %>% 
           str_remove(., "clade") %>% 
           str_remove(., "house") %>% 
           str_replace(., "World", " World ")) %>% 
  separate(name, into = c("predictor", "level"), sep = "_") %>% 
  mutate(predictor = if_else(predictor == "a", "predictor: clade", "predictor: house")) %>% 
  
  ggplot(aes(x = value, y = reorder(level, value))) +  # note how we used `reorder()` to arrange the coefficients
  geom_vline(xintercept = 0, color = "firebrick4", alpha = 1/10) +
  stat_pointinterval(point_interval = mode_hdi, .width = .89, 
                     size = 1, color = "firebrick4") +
  labs(x = "expected kcal (std)", 
       y = NULL) +
  theme_bw() +
  theme(axis.text.y = element_text(hjust = 0),
        axis.ticks.y = element_blank(),
        panel.grid = element_blank(),
        strip.background = element_blank()) +
  facet_wrap(~ predictor, scales = "free_y")
```

